Modelling the Real Time State of a Public Transport Road Network

### 2017-04-24 realtime-filter/realtime-filter.Rnw
writing algorithm and stuff

\documentclass[9pt]{IEEEtran}
%\documentclass[draftcls,a4paper,onecolumn]{IEEEtran}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{bbm}
\usepackage{subfig}

@@

Need to estimate vehicle state at time $t_k$, $\bX_k = [d_k, v_k, \ldots]$, 
from observations, $\bY_k = [\phi_k, \lambda_k]$.
Use a transition function to predict,
likelihood function to weight and update (via weighted resample).

@@

Algorithm~1 describes the particle filter transition algorithm that
describes the motion of a bus between observations.

@@

\begin{enumerate}
%\subsubsection[L1]{$L_1$: not at a stop or intersection}
\item
In situations where the bus is neither at a bus stop nor at an intersection,
the likelihood is simply determined from the distance between the particle and 
vehicle's reported position.
The particle filter framework makes this easy,
as each particle has a single value of distance into trip, $d_k^{(i)}$,
with which we can compute the latitude and longitude values based on 
the routes shape information, $z_k^{(i)} = h(d_k{(i)})$.

@@

The second part of the likelihood involves the stop time update,
and how the likelihood is formed depends on what data is available,
where $\bT_k = \left[ j, T^a_j , T^d_j \right]^T$ 
are the stop index, arrival and departure times,
respectively, either of which can be NA.

\begin{enumerate}
\item 
  If both $T^a_j$ and $T^d_j$ are NA, 
  then there is no additional likelihood;
  $ \log L_2  (\bX_k^{(i)} | \bT_k) = 0 $.

\item
  If departure time, $T^d_j$, is available only, for stop $j$, then we
  know the bus has at least departed stop $j$; therefore,
  \begin{equation}
    \label{eq:l2}
      \log L_2 (\bX_k^{(i)} | \bT_k) =
      \begin{cases}
        - \log(2\pi\sigma_d^2) - \frac{1}{2\sigma_d^2}(T^d_j - D_j^{(i)})^2 \\
        \hfill\quad \text{if } s_k^{(i)} \geq j \text{ and $D_j^{(i)}$ is set} \\
        - \infty \\
        \hfill\quad \text{otherwise}
      \end{cases}
  \end{equation}

\item
  If arrival time, $T^a_j$, is available only, then similarly
  \begin{equation}
    \label{eq:l2b}
    \log L_2  (\bX_k^{(i)} | \bT_k) =
    \begin{cases}
      - \log(2\pi\sigma_d^2) - \frac{(T^a_j - A_j^{(i)})^2}{2\sigma_d^2} & s_k^{(i)} \geq j \\
      - \infty & \text{otherwise}
    \end{cases}.
  \end{equation}
  Note that this does not assume the particle hasn't yet left the stop;
  the departure time could simply be missing data.

\item
  If both arrival time and departure time are available, then
  \begin{equation}
    \label{eq:l2c}
    \log L_2  (\bX_k^{(i)} | \bT_k) =
    \begin{cases}
      \begin{array}
        - \log(2\pi\sigma_d^2) - \frac{(T^a_j - A_j^{(i)})^2}{2\sigma_d^2} \\
        + \log(\|\bT_k\|) - \|\bT_k\|
        (D_k^{(i)} - A_k^{(i)})
      \end{array} \\
      \hfill\quad \text{if } s_k^{(i)} \geq j \text{ and $D_j^{(i)}$ is set}  \\
      - \infty\\
      \hfill\quad \text{otherwise}
    \end{cases}
  \end{equation}
  which makes use of the observed dwell time at the stop.

@@

@@

@@

\begin{itemize}
\item removing ``partial'' segments at start/end of routes
\end{itemize}

>>>>

%\documentclass[9pt]{IEEEtran}
\documentclass[draftcls,a4paper,onecolumn]{IEEEtran}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{bbm}
\usepackage{subfig}
\usepackage{algpseudocode}
\algblockdefx[Phase]{Phase}{EndPhase}[2]{\textbf{Phase #1:} #2}{\textbf{End Phase}\\}

@@

The primary goal of the particle filter is to estimate the unobservatble
vehicle state,
consisting of distance into trip ($d$), velocity ($v$),
current stop sequence ($s$), 
arrival and departure times at the last stop ($A_s$ and $D_s$),
current segment index ($r$),
the time spent queuing at the beginning of the current segment ($Q_r$),
and the time at which the vehicle began traveling along the current segment ($B_r$);
the vehicle state at time $k$ is denoted
\begin{equation}
  \label{eq:vehiclestate}
  \bX_k = [d_k, v_k, s_k, A_{s_k}, D_{s_k}, r_k, Q_{r_k}, B_{r_k}]^T.
\end{equation}

The vehicle state is related to an observable state $\bY_k = [\phi_k, \lambda_k]$,
(which contains the latitude and longitude, respectively)
through a \emph{measurment function}, $\bY_k = h(\bX_k) = h(d_k)$,
which depends only on the distance into trip.

There is a second source of data, the \emph{stop time updates}:
\mbox{$\bT_k = \left[ j, T^a_j , T^d_j \right]^T$}
are the stop index, arrival and departure times,
respectively; 
either of the arrival or departure times can be missing.

The particle filter is used to model a transit vehicle,
so we need to know about the behaviour of these.
The vehicle itself has a purpose of driving along an \emph{a priori} known route,
stopping at stops with known location along the way to pick up
and drop off passengers, if there are any.
Between stops, vehicles travel as close to the posted speed limit as possible,
typically only stopping at intersections (if necessary).
Due to many conditional situations, 
the particle filter becomes a vastly superior model when compared to traditional
Kalman filter based methods, as it can easily encompass each of these plausible states.

On approach to a stop $j$, the bus will stop with probability $\pi_j$ 
(i.e., it is a bernoulli trial for each particle).
If it stops, it will wait a minimum of $\gamma$ seconds while the doors are opened and closed,
and an additional $\bar t$ seconds as passengers board and disembark.
$\gamma$ is considered a fixed parameter, across all busses, 
while $\bar t$ is given an exponential distribution that, for now, has a fixed mean, $\tau$
(section~\ref{sec:discussion} discusses modifications to this).

On approach to an intersection (the end of a segment),
the behaviour is slightly different;
the vehicle doesn't reach the intersection itself, instead it joins the queue
(if there is one), and queues until it is able to pass through.
Similarly to stops, currently this is just modelled using fixed parameter values;
$\rho$ is the probability of stopping at the intersection
(for example, green vs.\ red traffic lights),
and $\nu$ is the average queue time.

To implement the particle filter model, 
we simply initialise $N$ particles (using superscript $^{(i)}$ to identify an individual particle)
by sampling distance and velocity, and then determining $s_k$ and $r_k$ from the 
vectors of stop and segment distances, $\bS^d$ and $\bR^d$, respectively.
The remaining state parameters are NULL.
Each particle is then passed through the \emph{transition function} independently (the prediction step),
and then the vehicle state estimate is updated using the data by computing likelihood weights
and resampling the particles with replacement.
The steps are described below.

@@

Algorithm~1 describes the particle filter transition algorithm that
describes the motion of a bus between observations.
The $k$ subscripts and $(i)$ superscripts have been dropped from the
vehicle state parameters (e.g., $d \equiv d_k^{(i)}$).

The additional parameters are the noise terms, $\sigma_v^2$ and $\sigma_r^2$,
which represent system noise and between-segment speed variance respectively.
The former is assumed to be small, and represents the average change in speed
of a bus per second \emph{within a segment},
while the latter is the variance in speed \emph{between adjacent segments},
and allows for the bus to quickly change speed when passing through an intersection.

\begin{figure}[bt]
  \begin{algorithmic}
    \Function{transition}{$\bX$, $\delta$, $t$}

    \Phase{1}{initial wait time}
    \State $w\gets 0$
    \If {$A_s \neq$ NULL $\cap\ D_s =$ NULL}
      \Comment{at stop}
      \State $w\gets w + \max\{0, \gamma - (t - A_s)\}$
      \Comment{min dwell time}
      \State $\bar t \sim \mathcal{E}(\tau^{-1})$
      \State $w\gets w + \bar t, \delta\gets \delta - w$
      \If {$\delta > 0$}
        \State $D_s\gets t - \delta$
        \Comment{depart stop}
      \EndIf
    \ElsIf {$Q_r \neq$ NULL $\cap\ B_r =$ NULL}
      \Comment{at intersection}
      \State $\bar t \sim \mathcal{E}(\nu^{-1})$
      \State $Q_r\gets Q_r + \bar t$, $w\gets w+\bar t$, $\delta\gets \delta - w$
      \If {$\delta > 0$}
        \State $B_s\gets t - \delta$
        \Comment{pass through intersection}
      \EndIf
    \EndIf
    \EndPhase

    \Phase{2}{System Noise}
    \State $v_k^{(i)}\sim \mathcal{N}(v, \delta\sigma_v^2)$
    \Comment{add system noise}
    \EndPhase

    \Phase{3}{Transition Forward}
    \State see figure~\ref{alg:transitionPhase3}
    \EndPhase

    \State \textbf{return} $\bX$
    \EndFunction
  \end{algorithmic}
  \caption{Particle filter transition function. Arguments are vehicle state~($\bX$), time since previous observation~($\delta$), and time of current observation ($t$).}
  \label{alg:transition}
\end{figure}

\begin{figure}[bt]
  \begin{algorithmic}
    \Phase{3}{Transition Forward}
    \While {$\delta > 0$}
      \If {$S^d_{s+1} \leq R^d_{r+1}$}
        \Comment{approaching stop}
        \State ETA  $\gets \frac{1}{v} (S^d_{s+1} - d)$
        \If {ETA $\leq \delta$}
        \Comment{reaches stop}
          \State $\delta\gets \delta - $ ETA, $s \gets s + 1$
          \State $A \gets t - \delta$, $D\gets $ null, $d\gets S^d_s$
          \If {$s = M$}
            \State\textbf{exit}
            \Comment{reached last stop}
          \EndIf
          \State $p\gets \mathcal{B}(\pi)$, $\bar t \sim \mathcal{E}(\tau^{-1})$
          \State dwell $\gets p(\gamma + \bar t)$, $\delta\gets \delta\ - $ dwell
          \If {$\delta > 0$}
            \State $D\gets t - \delta$
            \Comment{departs stop}
          \EndIf
        \Else
          \Comment{doesn't reach stop}
          \State $d\gets d + \delta v$, $\delta \gets 0$
        \EndIf
      \Else
        \Comment{approaching intersection}
        \State ETA $\gets \frac{1}{v}(R^d_{r+1} - d)$
        \If {ETA $\leq \delta$}
          \Comment{reaches intersection}
          \State $\delta\gets \delta - $ ETA, $r\gets r + 1$
          \State $Q \gets 0$, $B\gets $ null, $d\gets R^d_r$
          \State $p \sim \mathcal{B}(\rho)$, $\bar t \sim \mathcal{E}(\nu^{-1})$
          \State queue $\gets p\bar t$, $\delta\gets \delta\ - $ queue
          \State $Q\gets Q\ +$ queue
          \If {$\delta > 0$}
            \Comment{makes it through intersection}
            \State $B\gets t - \delta$
            \State $v \sim \mathcal{N}(v, \sigma_r^2)$
          \EndIf
        \Else
          \Comment{doesn't reach intersection}
          \State $d\gets d + \delta v$, $\delta \gets 0$
        \EndIf
      \EndIf
    \EndWhile
    \EndPhase
  \end{algorithmic}
  \caption{Transition function, phase 3: the main phase in which particles are moved forwards along the route.}
  \label{alg:transitionPhase3}
\end{figure}

@@

\begin{enumerate}
%\subsubsection[L1]{$L_1$: not at a stop or intersection}
\item
In situations where the bus is neither at a bus stop nor at an intersection,
the likelihood is simply determined from the distance between the particle and 
vehicle's reported position.
The particle filter framework makes this easy,
as each particle has a single value of distance into trip, $d_k^{(i)}$,
with which we can compute the latitude and longitude values based on 
the routes shape information, $z_k^{(i)} = h(d_k^{(i)})$.
The function $g (\bZ | \bY)$ performs an \emph{equirectangular projection} \cite{Snyder1997}
of the coordinates $\bX$, centered on $\bY$, 
onto a flat surface so that the Euclidean distance between points is equivalent
to the Great Circle distance between their associated coordinates.
This allows us to use a bivariate normal likelihood on the particles' positions,
with a GPS error of $\sigma_y^2$.

@@

The second part of the likelihood involves the stop time update,
and how the likelihood is formed depends on what data is available,
where $\bT_k = \left[ j, T^a_j , T^d_j \right]^T$ 
are the stop index, arrival and departure times,
respectively, either of which can be NULL.

\begin{enumerate}
\item 
  If both $T^a_j$ and $T^d_j$ are NULL, 
  then there is no change to the likelihood;
  $ L_2  (\bX_k^{(i)} | \bT_k) = 1 $.

\item
  If departure time, $T^d_j$, is available only, for stop $j$, then we
  know the bus has at least departed stop $j$; therefore,
  \begin{equation}
    \label{eq:l2}
    L_2 (\bX_k^{(i)} | \bT_k) =
      \begin{cases}
        f_N(D_j^{(i)} | T_j^d, \sigma_d^2) & s_k^{(i)} \geq j \cap D_j^{(i)} \neq \text{NULL} \\
        0 & \text{otherwise},
      \end{cases}
  \end{equation}
  where $f_N(x | \mu, \sigma^2)$ is the probability density of a normal distribution
  with mean $\mu$ and variance $\sigma^2$.

\item
  If arrival time, $T^a_j$, is available only, then similarly
  \begin{equation}
    \label{eq:l2b}
    L_2  (\bX_k^{(i)} | \bT_k) =
    \begin{cases}
      f_N(A_j^{(i)} | T_j^a, \sigma_d^2) & s_k^{(i)} \geq j \\
      0 & \text{otherwise}.
    \end{cases}
  \end{equation}
  Note that this does not assume the particle hasn't yet left the stop;
  the departure time could simply be missing data.

\item
  If both arrival time and departure time are available, then
  \begin{equation}
    \label{eq:l2c}
    L_2  (\bX_k^{(i)} | \bT_k) =
    \begin{cases}
      \begin{array}{l}
        f_N(A_j^{(i)} | T_j^a, \sigma_d^2) f_E(D_k^{(i)} - A_k^{(i)} | T_k^\delta) \quad\\
        \hfill s_k^{(i)} \geq j \cap D_j^{(i)} \neq \text{NULL} \\[0.5em]
        0 \qquad\qquad \text{otherwise},
      \end{array}
    \end{cases}
  \end{equation}
  where $T_k^\delta = T_k^d - T_k^a$ is the observed dwell time at the stop,
  and $f_E(x | \lambda)$ is the probability density of an exponential distribution 
  with rate $\lambda$.

@@

We now determine the likelihood of each particle by combining $L_1$ and $L_2$;
\begin{equation}
  \label{eq:particle_likelihood}
  L (\bX_k^{(i)} | \bY_k, \bT_k) = L_1(\bX_k^{(i)}|\bY_k) L_2(\bX_k^{(i)} | \bT_k).
\end{equation}
This is then used to compute the weight of each particle,
\begin{equation}
  \label{eq:particle_weights}
  w_i = \frac{L (\bX_k^{(i)} | \bY_k, \bT_k)}{\sum_{j=1}^N L (\bX_k^{(j)} | \bY_k, \bT_k)}.
\end{equation}

The resampling stage takes place at the beginning of each iteration for computational 
reasons (see section~\ref{sec:results}).
so we therefore use the weighted estimates of parameters in the next section.
However, the resampling is simply done using a standard weighted resampling 
algorithm with replacement.

@@

- mention why weighted estimates stored rather than resampled 
(i.e., to keep trajectories of disbanded particles for analysis).

@@

\begin{itemize}
\item removing ``partial'' segments at start/end of routes
\item enforcing max length of segments
\end{itemize}
